When does vapor pressure deficit drive or reduce evapotranspiration?

### 2019-02-28 doc/paper/body.tex
done with reviewer 3, now just edit paper make it nice and then response to editor

\begin{table}
  \caption{Definition of symbols and variables, with citation for how
    values are calculated, if applicable.}
  \label{definitions}
  \centering \footnotesize
  \begin{tabular}{l c c c}
    \hline
    Variable & Description & Units & Citation \\
    \hline
    $e_s$  & saturation vapor pressure & Pa  & - \\
    $T$  & temperature  & K & - \\
    $P$  & pressure & Pa  & - \\
    $\Delta$  & $\frac{\partial e_s}{\partial T}$ & Pa K$^{-1}$ & - \\
    $R_{net}$  & net radiation at land surface minus ground heat flux & W m$^{-2}$   & - \\
    $g_a$  & aerodynamic conductance & m s$^{-1}$  & \citet{Shuttleworth_2012} \\
    $\rho_a$  & air density & kg m$^{-3}$  & - \\
    $c_p$  & specific heat capacity of air at constant pressure & J K$^{-1}$ kg$^{-1}$ & - \\
    $VPD$  & vapor pressure deficit & Pa  & - \\
    $\gamma$  & psychometric constant & Pa K$^{-1}$   & - \\
    $g_{s-leaf}$  & leaf-scale stomatal conductance & m s$^{-1}$  & \citet{MEDLYN_2011} \\
    $g_{s}$  &  stomatal conductance & m s$^{-1}$
                                   & \citet{Medlyn_2017} \\
    $g_{1-leaf}$  & leaf-scale slope parameter & Pa$^{0.5}$
                                   & \citet{MEDLYN_2011} \\
    $g_{1}$  & ecosystem-scale slope parameter & Pa$^{0.5}$ & \citet{Medlyn_2017} \\
    $R$ & universal gas constant & J mol$^{-1}$ K$^{-1}$ & - \\
    $R_{air}$ & gas constant of air & J  K$^{-1}$ kg$^{-1}$ & - \\
    $\sigma$ & uncertainty parameter & -& - \\
    $c_a$ & CO$_2$ concentration & $\mu$ mol CO$_2$ mol$^{-1}$ air& - \\
    $\lambda$ & marginal water cost of leaf carbon & mol H$_2$O mol$^{-1}$ CO$_2$ & - \\
    $\Gamma$ & CO$_2$ compensation point & - & - \\
    $\Gamma^*$ & CO$_2$ compensation point without dark respiration & - & - \\
    \hline
  \end{tabular}
\end{table}

@@

  This derivation relied either implicitly or explicitly on several
  assumptions. First, we assume that VPD at the leaf surface is the
  same as VPD at measurement height; physically this implies that
  leaves are perfectly coupled to the atmosphere. In reality, for some
  conditions and plant types the leaves can become decoupled from the
  boundary layer \citep{De_2017, Medlyn_2017}. Therefore, our
  derivation will be most applicable in times like the growing season
  (when we also expect uWUE to be most valid), when relatively high
  insolation induces instability and convective boundary layers, and
  we would expect the surface to be generally well coupled. An
  additional assumption in the formulation of $uWUE$ \citep{Zhou_2014,
    Zhou_2015} and \citet{Medlyn_2017}'s stomatal conductance model is
  that direct soil evaporation (E) contributions to ET remain small
  relative to transpiration (T). Again, this should be more true
  during the growing season. The ratio of E to T may increase
  immediately after rainfall events due to high soil moisture and
  ponding, but VPD is generally low anyways during these
  times. However, some plant types allow for systematically larger
  contributions of E in ET, particularly those with sparse canopies
  and smaller relative amounts of transpiration. We therefore might
  expect that the theory will be most applicable to forest PFTs, which
  will be most strongly coupled to the boundary layer due to larger
  surface roughness, and will also generally have the highest ratios
  of transpiration to evaporation. Finally, we assume that $g_1$ and
  $uWUE$ are constant with respect to the conceptual VPD
  perturbation. Both quantities have been shown to be relatively
  constant with respect to changes in
  VPD \citep{Franks_2017, Zhou_2014}. These parameters will however
  vary with plant species and characteristics \citep[e.g. wood density,
  ][]{Lin_2015}, as well as environmental conditions including soil
  water content \citep{Manzoni2013} and temperature.

>>>>

\begin{table}
  \caption{Definition of symbols and variables, with citation for how
    values are calculated, if applicable.}
  \label{definitions}
  \centering \footnotesize
  \begin{tabular}{l c c c}
    \hline
    Variable & Description & Units & Citation \\
    \hline
    $e_s$  & saturation vapor pressure & Pa  & - \\
    $T$  & temperature  & K & - \\
    $P$  & pressure & Pa  & - \\
    $\Delta$  & $\frac{\partial e_s}{\partial T}$ & Pa K$^{-1}$ & - \\
    $R_{net}$  & net radiation at land surface minus ground heat flux & W m$^{-2}$   & - \\
    $g_a$  & aerodynamic conductance & m s$^{-1}$  & \citet{Shuttleworth_2012} \\
    $\rho_a$  & air density & kg m$^{-3}$  & - \\
    $c_p$  & specific heat capacity of air at constant pressure & J K$^{-1}$ kg$^{-1}$ & - \\
    $VPD$  & vapor pressure deficit & Pa  & - \\
    $\gamma$  & psychometric constant & Pa K$^{-1}$   & - \\
    $g_{0}$  & leaf-scale mimimum stomatal conductance & m s$^{-1}$  & \citet{MEDLYN_2011} \\

    $g_{s-leaf}$  & leaf-scale stomatal conductance & m s$^{-1}$  & \citet{MEDLYN_2011} \\
    $g_{s}$  &  stomatal conductance & m s$^{-1}$
                                   & \citet{Medlyn_2017} \\
    $g_{1-leaf}$  & leaf-scale slope parameter & Pa$^{0.5}$
                                   & \citet{MEDLYN_2011} \\
    $g_{1}$  & ecosystem-scale slope parameter & Pa$^{0.5}$ & \citet{Medlyn_2017} \\
    $R$ & universal gas constant & J mol$^{-1}$ K$^{-1}$ & - \\
    $R_{air}$ & gas constant of air & J  K$^{-1}$ kg$^{-1}$ & - \\
    $\sigma$ & uncertainty parameter & -& - \\
    $c_a$ & CO$_2$ concentration & $\mu$ mol CO$_2$ mol$^{-1}$ air& - \\
    $\lambda = \frac{\partial \; transpiration}{\partial\; CO_2\; assimilation}$ & marginal water cost of leaf carbon & mol H$_2$O mol$^{-1}$ CO$_2$ & - \\
    $\Gamma$ & CO$_2$ compensation point & - & - \\
    $\Gamma^*$ & CO$_2$ compensation point without dark respiration & - & - \\
    \hline
  \end{tabular}
\end{table}

@@

  This derivation relied either implicitly or explicitly on several
  assumptions. First, we assume that VPD at the leaf surface is the
  same as VPD at measurement height; physically this implies that
  leaves are perfectly coupled to the atmosphere. In reality, for some
  conditions and plant types the leaves can become decoupled from the
  boundary layer \citep{De_2017, Medlyn_2017}. Therefore, our
  derivation will be most applicable in times like the growing season
  (when we also expect uWUE to be most valid), when relatively high
  insolation induces instability and convective boundary layers, and
  we would expect the surface to be generally well coupled. An
  additional assumption in the formulation of $uWUE$ \citep{Zhou_2014,
    Zhou_2015} and \citet{Medlyn_2017}'s stomatal conductance model is
  that direct soil evaporation contributions to ET remain small
  relative to transpiration. Again, this should be more true during
  the growing season. The ratio of evaporation to transpiration may
  increase immediately after rainfall events due to high soil moisture
  and ponding, but VPD is generally low anyways during these
  times. However, some plant types allow for systematically larger
  contributions of evaporation in ET, particularly those with sparse
  canopies and smaller relative amounts of transpiration. We therefore
  might expect that the theory will be most applicable to forest PFTs,
  which will be most strongly coupled to the boundary layer due to
  larger surface roughness, and will also generally have the highest
  ratios of transpiration to evaporation. Finally, we assume that
  $g_1$ and $uWUE$ are constant with respect to the conceptual VPD
  perturbation. Both quantities have been shown to be relatively
  constant with respect to changes in VPD \citep{Franks_2017,
    Zhou_2014}. These parameters will however vary with plant species
  and characteristics \citep[e.g. wood density, ][]{Lin_2015}, as well
  as environmental conditions including soil water content
  \citep{Manzoni2013} and temperature.

### 2019-02-28 doc/paper/hess-revision/final-response.tex
done with reviewer 3, now just edit paper make it nice and then response to editor

\subsection{(ii)}

As in our
\href{https://www.hydrol-earth-syst-sci-discuss.net/hess-2018-553/hess-2018-553-AC2-supplement.pdf}{previous
  response}, we agree that we need more discussion connecting and
contrasting leaf level results with our ecosystems scale approach. We
have summarized the discussion from our reviewer response and added it
to the introduction in lines ** **.

@@

We agree that are PFT-based formulation was too easy to be
misintrepreted as an assertion that g1 and uWUE are constant and never
vary in time or space. As described in Section \ref{pft}, we have
removed the observational and PFT-focused component of the analysis,
and instead focus exclusvively on using previously published constants
(and their range) to interpret results. We would also like to
reiterate that we do not view our use of uWUE as an introduction of an
extra parameter. Our use of uWUE replaces a photosynthesis model,
which will generally have many more parameters. Using uWUE is instead
a simplifying approximation that relies on assumptions (based on
optimal theory), but one that certainly reduces the total number of
parameters of a more traditional estimate of ET.

@@

  the sufficient understanding, and included explicit discussion of

@@

  \citep{Lin_2005, Franks_2017, Zhou_2015}, so would be interested in
  citations showing that they are not constant for the majority fo
  conditions. Please note we also added some plots of the relationship
  between soil moisture and uWUE in our supplementary material to aid
  the reader, but view this as an active research as well.

@@

@@

>>>>

\subsection{(ii)}
\label{leaf}
As in our
\href{https://www.hydrol-earth-syst-sci-discuss.net/hess-2018-553/hess-2018-553-AC2-supplement.pdf}{previous
  response}, we agree that we need more discussion connecting and
contrasting leaf level results with our ecosystems scale approach. We
have summarized the discussion from our reviewer response and added it
to the introduction in lines ** **.

@@

We agree that are PFT-based formulation was too easy to be
misintrepreted as an assertion that g1 and uWUE are constant and never
vary in time or space. As described in Section \ref{pft}, we have
removed the observational and PFT-focused component of the analysis,
and instead focus exclusvively on using previously published constants
(and their range) to interpret results. We would also like to
reiterate that we do not view our use of uWUE as an introduction of an
extra parameter. Our use of uWUE replaces a photosynthesis model and
stomatal conductance reduction factor, which will generally have many
more parameters. Using uWUE is instead a simplifying approximation
that relies on assumptions (based on optimal theory), but one that
certainly reduces the total number of parameters used in a more
traditional estimate of ET.

@@

  have sufficient understanding, and included explicit discussion of

@@

  \citep{Lin_2005, Franks_2017, Zhou_2015}, so we would be interested
  in citations showing that they are not constant for the majority fo
  conditions. We also added some plots of the relationship between
  soil moisture and uWUE in our supplementary material to aid the
  reader, but view those as inconclusive and an active research as
  well.

@@

\item P 23 LL 16-21 : we added language about the leaf-scale derived
  shape of the ET-VPD curve in the introduction. We prefer this
  section in the discussion, as we view it as a discussion of the
  implication of our results (which is the derivation itself), and how
  it is sensitive to stomatal model choice.

@@

\section{Reviewer 3}

Thank you for a very high quality review.

\subsection{Consequences of $\lambda$ in g$_1$ and uWUE}

\subsubsection{Soil moisture and $\lambda$}

Yes, $\lambda$, and by extension uWUE and g$_1$ can vary with
environental conditions, including soil moisture. In the worst case,
the consequences of this dependency for scientists applying our theory
is that they must conceptually consider the change in ET with respect
to VPD only for scenarios in which soil moisture is fixed. We believe
this more restricted scenario still has utility for disentangling land
atmosphere feedbacks, and understanding observations and more
sophisticated models. However, \cite{Zhou_2015}'s results do suggest
that uWUE (and by extension $\lambda$) may be somewhat constant with
respect to soil moisture. The degree to which it is constant
influences the degree to which a reader can relax their interpretation
to one in which soil moisture is not fixed with respect to VPD. We
attempted to shed some more light on this issue by plotting the
distribution of uWUE with SWC for 66 fluxnet sites, but the signal to
noise ratio was too low and when there was a signal of a relationship,
it was inconsistent across sites. We did include this in supplemental
material to aid the reader and draw attention to uWUE (and $\lambda$)
- SWC relationships at the ecosystem scale as an important area for
future research.

\subsubsection{Interpretation issues with the use of the uncertainty
  parameter}

We agree that forcing all uncertainty (and real variability) into a
single parameter made succesful interpretation our results extremely
difficult. The general issues with this formulation lead us to remove
the observational component of the paper, to focus on the derivation
using plant parameters previously established in obsevational studies
in the literature \citep{Zhou_2015, Medlyn_2017}. This simplification
and focus also [hopefully] clarifies the manuscript and the
presntation of results.

\subsection{General comments}

\begin{itemize}
  \item \textbf{Symbol definition}: thank you for pointing out these mistakes
    and/or potential sources of confusion. We have gone through the
    manuscript and attempted to clarify/fix these issues.
  \item \textbf{Dependence on $\lambda$}: we did present both uWUE and g$_1$ in
    terms of $\lambda$ in our original manuscript, but we added
    language drawing the readers attention to this, and elaborating
    its importance for interpreting how they co-vary.
  \item \textbf{Presentation of leaf-level stomatal conductace}: we thought
    this might help the reader and draw attention that it the theory
    originated at the leaf scale. However we agree with you that it
    seems unnecesary given we do not use it, and makes the manuscript
    less clear. We removed it.
  \item \textbf{g$_a$ calculation}: in the newer idealized version of
    the manuscript we do not calculate g$_a$, but represent a range of
    reasonable values suggested by data and the literature
    \citep[e.g.,][]{Raupach_1995}.
  \item \textbf{Fitting of parameters with FLUXNET data}: as
    mentioned, we removed this part of the analysis.
\end{itemize}

\subsection{Presentation of results}

We agree that the results are more clear when all figures are in terms
of $\frac{\partial ET}{\partial VPD}$. All figures are now presented
in terms of $\frac{\partial ET}{\partial VPD}$, including those where
only parameters in the sign term are varied.

We also agree that our original manuscript was confusing and
disjointed. Hopefully the signficiant changes to the structure of the
manuscript have clarified, streamlined, and simplifed the
results. Regarding Section 3.7, we moved it into a new ``discussion''
section. While it could be placed in the methods, we view it as more
of a discussion on the consequences of using a given stomatal
conductance theory, and think it is hopefully easier for the reader to
digest after seeing the concave up nature of our results.

Yes, we did misuse ``leading order'' language, which we removed. Also many
of the subjective interpretations of quantiative results have been
removed with the removal of the observational component.

\subsection{Minor comments}

\begin{itemize}
  \item \textbf{Introduction}: We added more discussion of previous
    work on transpiration response to VPD, as elaborated on in Section
    \ref{leaf}. We also introduce \cite{Novick_2016}'s results, and
    added langauge recognizing there are more details to plant strategy
    and response to VPD than our simplified logic in the introduction.
  \item \textbf{P 9 results equations}: These were moved to methods.
  \item \textbf{P 13, L32, Temperature effect}: Yes, the direct effect is
    included in our idealized analysis (although because of CC the
    effect through $\Delta$ should be larger than the ``direct''
    effect).
  \item \textbf{P 15 L 15 ``between''}: removed.
\end{itemize}

### 2019-02-28 doc/paper/intro.tex
done with reviewer 3, now just edit paper make it nice and then response to editor

The ET response to changes in VPD alters water partitioning between
the soil and atmosphere. If ecosystem plant response reduces ET with
atmospheric drying then soil moisture will be better conserved. This
represents a sensible evolutionary strategy to cope with aridity: save
water for periods when atmospheric demand for water is relatively low,
and atmospheric carbon can be accessed with a relatively smaller cost
in water loss. If instead stomata were fully passive \citep [similar
to soil pores, e.g. ][]{Or_2013}, increased atmospheric aridity would
strongly reduce soil moisture \citep{Berg_2017}. This could further
increase aridity as low soil moisture levels increases the Bowen
ratio, leading to increased temperature and atmospheric drying
\citep[][]{Bouchet_1963, Morton_1965, Brutsaert_1999, Ozdogan_2006,
Salvucci_2013, Gentine_2016, Berg_2016, Zhou_2019}. Therefore, passive
regulation and a lack of soil moisture conservation does not seem to
be a sensible strategy for plants from an evolutionary standpoint.

** paragraph on soil moisture dependenec? **

@@

In this manuscript, we leverage uWUE and recent developments in
stomatal conductance parameterizations \citep{MEDLYN_2011} with a
Penman-Monteith framework \citep [hereafter PM,][]{Penman_1948,
Monteith_1965} to derive the theoretical one-way response of ET to VPD
with other environmental variables properly controlled for, i.e. we
develop a framework for evaluating the partial derivative of ET with
respect to VPD. We explicitly include VPD's full effect on stomatal
conductance, including its impact on photosynthesis. The use of an
energy balance framework (PM) allows us to include in our analysis the
effects of the energy cost of evaporating water from a surface, which
is an important factor in the natural environment, compared to
prescribed in situ envirnmental conditions. This is relevant because
previous research focusing on the leaf scale \citep{Rawson1977,
Turner1984, Oren1999, Darmour2010, Mott2013} does not consistently or
analogously include the energetic constraints on evapotranspiration
under which ecosystems operate. The leaf-scale results agree that
stomatal conductance decreases in response to VPD \citep{Oren199,
Damour2010}, which we expect to be true for an ecosystem as
well. However leaf scale results also indicate that transpiration
usually increases with increasing VPD in a concave downward shape
\citep[e.g.,][]{Rawson1977, Turner1983, Mott2013}, which may not be
true for ET at the ecosystem scale once energetic constraints on ET
are included in the analysis. Our approach allows us to estimate the
expected ET response to VPD at the ecosystem scale, including the
effects of surface energy balance constraints, and assess how
ecosystem ET response to VPD deviates from previous leaf scale
analyses. It is indeed key to disconnect the impact of VPD from other
variables as VPD is known to increase dramatically with future climate
** (Novick JGR 2017) **, yet other environmental conditions such as
soil moisture and radiation will not increase as
dramatically. Nonetheless previous studies have typically been
hampered by the covariations between different variables
\citep[e.g. radiation and VPD or soil moisture and VPD, ][]{Lin_2018,
Zhou_2019}, which limits the interpretability and understanding of the
response of ecosystem-scale ET to VPD.

>>>>

The ET response to changes in VPD alters water partitioning between
the soil and atmosphere. If ecosystem plant response reduces ET with
atmospheric drying then soil moisture will be better conserved. This
represents a sensible evolutionary strategy to cope with aridity: save
water for periods when atmospheric demand for water is relatively low,
and atmospheric carbon can be accessed with a relatively smaller cost
in water loss. If instead stomata were fully passive \citep [similar
to soil pores, e.g. ][]{Or_2013}, increased atmospheric aridity would
strongly reduce soil moisture \citep{Berg_2017}. This could further
increase aridity as low soil moisture levels increases the Bowen
ratio, leading to increased temperature and atmospheric drying
\citep[][]{Bouchet_1963, Morton_1965, Brutsaert_1999, Ozdogan_2006,
Salvucci_2013, Gentine_2016, Berg_2016, Zhou_2019}. Therefore, passive
regulation and a lack of soil moisture conservation does not seem to
be a sensible strategy for plants from an evolutionary
standpoint. This simplified logic explains generally why plants
evolved to respod to VPD, but also excludes many possible details and
special cases (e.g. plant to plant ineractionand highly specialized
photosythesis strategies like Crassulacean acid metabolism
photosntheisis).

@@

In this manuscript, we leverage uWUE and recent developments in
stomatal conductance parameterizations \citep{MEDLYN_2011} with a
Penman-Monteith framework \citep [hereafter PM,][]{Penman_1948,
Monteith_1965} to derive the theoretical one-way response of ET to VPD
with other environmental variables properly controlled for, i.e. we
develop a framework for evaluating the partial derivative of ET with
respect to VPD. It is indeed key to disconnect the impact of VPD from
other variables as VPD is known to increase dramatically with future
climate and limit ET more than soil moisture
\citep{Novick_2016}. Nonetheless previous studies have typically been
hampered by the covariations between different variables
\citep[e.g. soil moisture and VPD, ][]{Lin_2018, Zhou_2019}, which
limits the interpretability and understanding of the response of
ecosystem-scale ET to VPD. We are able to disentangle the effect of
VPD on ET because we explictly include VPD's full effect on stomatal
conductance, including its impact on photosynthesis. The use of an
energy balance framework (PM) allows us to include in our analysis the
effects of the energy cost of evaporating water from a surface, which
is an important factor in the natural environment, compared to
prescribed in situ envirnmental conditions. This is relevant because
previous research focusing on the leaf scale \citep{Rawson1977,
Turner1984, Oren1999, Darmour2010, Mott2013} does not consistently or
analogously include the energetic constraints on evapotranspiration
under which ecosystems operate. The leaf-scale results agree that
stomatal conductance decreases in response to VPD \citep{Oren1990,
Damour2010}, which we expect to be true for an ecosystem as
well. However leaf scale results also indicate that transpiration
usually increases with increasing VPD in a concave downward shape
\citep[e.g.,][]{Rawson1977, Turner1983, Mott2013}, which may not be
true for ET at the ecosystem scale once energetic constraints on ET
are included in the analysis.  Our approach allows us to estimate the
expected ET response to VPD at the ecosystem scale, including the
effects of surface energy balance constraints, and assess how
ecosystem ET response to VPD deviates from previous leaf scale
analyses.
